### 2016-05-03 article_ru.tex

%We are specifying language of the paper
\selectlanguage{english}
\def\dd#1#2{\frac{\partial#1}{\partial#2}}
\section{
%%%%%%%%%%%%%%% You have to specify name of your paper here
``Investigation of tuning parameters of tabu search algorithm and its modification  for solving the static Routing Courier Delivery Problem.'' 
}

\authors{
R.~Shafeyev , L.~Lyubchik 
}

@@

The paper considers the routing courier delivery problem with the service time for which the discrete model was constructed and the calculation scheme based on tabu search algorithm. The efficiency  of the proposed algorithm‘s scheme has been tested on a large-scale problems, which were generated on the basis of different classes known model routing problems.

%%%%%%%%%%%%%%% To define new section use a command \subsection
% We are defining "Introduction" here
\subsection{Introduction}
%%%%% Paragraph text. To separate paragraphs use empty lines.
In the research we consider a static vehicle routing problem with time windows. This variant of the routing problem is a common problem in the courier delivery industry. Effective solution to this problem affects the quality of services provided by the company, as well as the magnitude of potential losses caused by the excess of the agreed terms of delivery. The proposed solution to the problem is based on the automation of the process of planning the route and involves the use of distributed information systems for the possible implementation of these approaches in practice. Another important criterion for good transport plan are low transport costs, which affect the cost of services.

The Routing Courier Delivery Problem with time windows requires that a set of vehicles satisfy a collection of customer requests. Each client request requires the use of a single vehicle to receive and load a specified amount of goods at one source location(sender) and to deliver them to destination location(receiver). All requests must be performed without violating either the vehicle capacity or the customer time window. The solution is a master plan that solves the problem, and takes into account the daily events that take place according to plan. In this paper, we present a discrete model and a modified tabu search algorithm  to solve the  Courier Delivery Problem with time windows.

In order to validate our strategies decisions technique effectiveness, we have constructed a new set of benchmark problems for the pickup and delivery problem with time windows based on which were generated on the basis of different classes known model routing problems. Computational results substantiate the solution quality and efficiency of our modified tabu search algorithm.

\subsection{Problem Statement}
Let $C,dim(C)=n$ is a set of vehicles and $Q,dim(Q)=m$ is a set of requests received from customers at the current time.

Suppose that the following information is known about vehicles from set $C$: \\
$\vec{P}_c$ -- vehicle position, $c \in C$;\\
$L_c$ -- vehicle capacity, $c \in C$.

Let $S$ is a set of senders, ($dim(S) = dim(Q) = m$), $R$ is a set of receivers, $dim(R) = dim(Q) = m$. Then each client request $q \in Q$ includes the following information:\\
$s_q$ -- a sender of client shipment, $s \in S$; \\
$r_q$ -- a receiver of client shipment, $r \in R$; \\
$\vec{P}_{s_q}$ -- sender position;\\
$\vec{P}_{r_q}$ -- receiver position; \\
$w_q$ -- client shipment that is required to deliver from the sender to the recipient; \\
$[t_{s}^{q}, t_{s}^{q}+ \Delta{t_{s}^{q}}]$ -- the time window within which worker must to pick up the goods from the sender;\\ 
$[t_{r}^{q}, t_{r}^{q}+ \Delta{t_{r}^{q}}]$ -- the time window within which worker must deliver the goods to the receiver.

Thus, each request can be represented as a tuple:
\begin{equation}
\forall q \in Q: q = (s_q,r_q, \vec{P}_{s_q}, \vec{P}_{r_q}, w_q, t_{s}^{q}, \Delta{t_{s}^{q}}, t_{r}^{q}, \Delta{t_{r}^{q}})
\end{equation}
To estimate the cost of transportation between destinations defined cost function $\Omega$:
\begin{equation}
\forall i,j \in S \cup R: \exists \Omega_{i,j} = \Omega(\vec{P}_i,\vec{P}_j)
\end{equation}

Necessary to construct the optimal routes of vehicles movement for the transportation of goods from the sender to the receiver for all client requests.

@@

\subsection{Discrete model}
Routing problem can be represented as a directed graph $G=G(V,E)$. The set $V=C\cup{S}\cup{R}$ -- nodes of the graph $G$, consisting of elements of vehicles, senders and receivers. $E$ -- dynamic set of arcs of the graph $G$, such that:
\begin{equation}
\forall e(X) \in E: e(X) = (v_i,v_j), \exists v_i \in V, v_j \in V/C
\end{equation}

Let $X = \{X^k\}^n_{k=1}$ is a sequence of variable matrices for each vehicle $k \in C$. Elements of the matrices take the following values:
\begin{equation}
  x^{(k)}_{i,j} = 
    \begin{cases}
	  1,&\text{the vehicle $k \in C$ moves from $i$ node to the $j$}\\
	  0,&\text{otherwise.}
    \end{cases}
\end{equation}
where: $i\in{V/C \cup {k}}, j \in V/C$.

Let us introduce the vector  of variables $ \vec{Y}  ^ k(X) $ for each vehicle $ k \ in C $. Vector elements have the following values:
\begin{equation}
\vec{y}^{(k)}_{j}(X) = 
\begin{cases}
1,&\text{the request $j \in Q$ is processed by vehicle $k \in C$}\\
0,&\text{otherwise.}
\end{cases}
\end{equation}
where: $i\in{V/C \cup {k}}, j \in V/C$.

Let $t^k_j(X^{(k)})$ -- arrival time of the vehicle $k \in C$ at the destination $j \in S \cup R$.

The objective function takes the following form:
\begin{equation} \label{main_objective}
  F(X) = 
    \sum_{k \in C}
     \sum_{i,j\cup{V}} 
     \Omega_{ij} \cdot x_{ij}^{(k)} 
     \to min
\end{equation}

We define constraints on the objective function (\ref{main_objective}), which provided the continuity of routes:
\begin{align} 
& \sum_{k \in C}\sum_{j \in S \cup R}x^{(k)}_{i,j} \leq 1, 
\forall i \in V \label{main_cond_1}\\
& \sum_{k \in C}\sum_{i \in S \cup R \cup \{k\} } x^{(k)}_{i,j} = 1, 
\forall j \in S \cup R \label{main_cond_2}\\
& \sum_{i \in S \cup R \cup \{k\} } x^{(k)}_{i,\omega} - 
\sum_{j \in S \cup R} x^{(k)}_{\omega,j} \leq 1, 
\forall \omega \in S \cup R,  \forall k \in C \label{main_cond_3}\\
&  \sum_{i \in S \cup R / Z}\sum_{j \in Z } x^{(k)}_{i,j} > 0, 
Z=\{z \in Z: \sum_{j \in S \cup R}x^{(k)}_{j,z}>0 \}  ,\forall k \in C \label{main_cond_4}
\end{align}

The restriction (\ref{main_cond_1}) prohibits to a node in the graph $G$ has more than one output arc. The restriction (\ref{main_cond_2}) prohibits to a node has more than one input arc.  The constraint (\ref{main_cond_3}) indicates that the number of input arcs to the node can not be less than the output arcs (this constraint considers the fact that the vehicle can leave the destination only if it has visited this node). The restriction (\ref{main_cond_4}) excludes local loops.

Then next constraints synchronize values of variables $X$ and $\vec{y}$ for each request $q \in Q$ and prohibits the service of receiver before a visit to the sender:
\begin{align} 
& x^{(k)}_{s_q} + x^{(k)}_{r_q} = 2 \cdot y^{(k)}_{q}(X), \forall k \in C, q \in Q  \\
& y^{(k)}_{q}(X) \cdot (\tilde{t}^k_{r_q}(X^{(k)})-\tilde{t}^k_{s_q}(X^{(k)})\ge{0}, \forall k \in C, q \in Q
\end{align}

Define restrictions for accounting of vehicle capacity and time windows:
\begin{align} 
& \sum_{j\in{Q}} \omega_j \cdot y_{j}^{k} \leq L_k, \forall{k}\in{C}\\
& t_{s}^{q} \leq \tilde{t}^k_{s_q}(X^{(k)} \leq t_{s}^{q}+ \Delta{t_{s}^{q}}, \forall q \in Q, \label{tws_cond} \\
& t_{r}^{q} \leq \tilde{t}^k_{r_q}(X^{(k)} \leq t_{r}^{q}+ \Delta{t_{r}^{q}}, \forall q \in Q, \label{twr_cond}
\end{align}

\subsection{Description of the Algorithm}
Present the set of matrices $\{X^k\}$ in the form of a vector defined on the hypercube $E^\eta=\{0.1\}^\eta$, where $\eta=n\cdot(2m+1)\cdot 2m$. The aim of the task is to minimize the objective function:
\begin{align} 
& F(\vec{u})\to min,\vec{u}\in E^{\eta}
\end{align}

Denote by $\delta(\vec{u},\vec{v})$ the Hamming distance between $\vec{u}$ and $\vec{v}$. Through $N_l(\vec{u})$ we denote the neighborhood of a point $\vec{u}$ radius $l$, ie \cite{vrp_shafeyev}:
\begin{align} 
& N_l(\vec{u})=\{\vec{v} \in E^{\eta}:\delta(\vec{u},\vec{v})\le l \}, l=\bar{1,\eta}
\end{align}

When $l=\eta$ a set $ N_l(\vec{u})$ for any vector $\vec{u}$ coincides with the set $E^{\eta}$ and being in this neighborhood vector with a minimum value of the objective function is equivalent to solving the original problem. The standard algorithm local search algorithm starts with a randomly selected vector $\vec{u^0}$.
On the $i$ step of the algorithm move from the current vector in the neighboring that the minimum value of the objective function in the neighborhood of the given vector:
\begin{align} 
& F(\vec{u}^{i+1})=min\{F(\vec{v}):\vec{v} \in N_l(\vec{u}^i)\}
\end{align}

The algorithm terminates at a local optimum, when $F(\vec{u^{i+1}})=F(\vec{u^i})$. When solving the transportation problem is a typical situation when many local optimums and only one of them is global:
\begin{align} 
& F_{opt}=min\{F(\vec{v}):v \in E^{\eta}\}
\end{align}

@@

\subsection{Initialization}
To use a search algorithm is restricted by the need to build the initial solution. To obtain the initial solution $\vec{u^0}$ you can use a heuristic method to construct the route. The essence of these methods is the following. Every car has consistently built a route by adding another is not considered clients according to specified rules.

The most known method of construction of the route proposed by Solomon in 1987 year and found in the literature as a method of Solomon. This algorithm is fast enough (computational complexity $O(n^3)$, therefore, it is useful to initialize a vector $\vec{u^0}$.

At the beginning of the algorithm, you add one to a customer on the route according to time limits. The choice of the first client is randomly or chosen one, you want to serve first. Then, each valid client $u$ from the set of outstanding clients $Q^{'}$ (the set $Q^{'}\subset{Q}$) evaluated insertion between two adjacent nodes route. For selection of inserts used criterion:
\begin{align} 
& c(v_i,u,v_i+1)=[\min_{p=2,\dots,n^k}(\Omega_{p-1,u}^k+\Omega_{u,p}^k-\mu \Omega_{p-1,p}^k) ]
\end{align}	
where:

$k$ - vehicle route $k\in C$;
$n^k$ - the number of nodes in the route;
$\mu$ - arbitrary parameter, $\mu \ge 0 $.

Added the client, which is $c(v_i,u,v_i+1)$ least:
\begin{align} 
& u^*:c(v_i,u^*,v_i+1)=min[c(v_i,u,V_i+1)]
\end{align}

As a result the client $u^*$ is added to the current route $k$ the position between nodes $v_i$ and $v_(i+1)$.

\subsection{A Method of Forming Solutions Neighborhood}
The most difficult part of the local search algorithm with the prohibitions are viewing the surroundings $N_l(\vec{u^i},\vec{\phi^i})$ of the current point $\vec{us^i}$. For even larger values of $l$ and the parameter $p=1$ in the outskirts plenty of points. But due to the nature of the model's constraints many of them are not the solution. Therefore, due to the nature of the considered discrete tasks, we used the following method of determining the area. On the $i$ step of the algorithm compiled from routes pair: $\{R_1,R_2\},\{R_1,R_3\},\dots,\{R_1,R_n\},\{R_2,R_1\}, \{R_2,R_3\},\dots,\{R_2,R_n\},\dots,\{R_{n-1},R_{n}\}$. Then over each pair of routes is performed in moving requests. The resulting displacement vectors $\vec{u}$ are neighborhood $N_l(\vec{u^i},\vec{\phi^i})$. For moving  we used change-move  and swap-move operations.

@@

As should have been expected, for large values of the parameter a jam occurs in local optimum due to the large number of restrictions of movement in space. Otherwise if you select a too small tabu size, algorithm can still get stuck in a local optimum. In the first computational experiment we made finding the best solutions for different values of the tabu size parameter from interval $h \in [0, 40]$ (pict. \ref{aba:fig2} an pict. \ref{aba:fig3}).
\begin{figure}[h]
	\hfil\includegraphics[height=2.2in]{images/tabuSize_time}\hfil
	\caption{
	  The dependence of the quality of solutions on the size of the tabu list when using $ TimeSpent $ termination strategy
	}
	\label{aba:fig3}
\end{figure}

>>>>

%We are specifying language of the paper
\selectlanguage{russian}
\def\dd#1#2{\frac{\partial#1}{\partial#2}}
\section{
``Исследование  настроечных параметров алгоритма поиска с запретами и его модификация для решения курьерской задачи маршрутизации.'' 
}

\authors{
L.~Lyubchik, R.~Shafeyev 
}

@@

В статье рассматривается задача маршрутизации курьерской доставки  с долгим сроком службы, для которой была построена дискретная модель и вычислительная схема, в основу которой был положен алгоритм поиска с запретами. Эффективность схемы предложенного алгоритма была проверена на задачах большой размерности, которые были составлены на базе известных модельных задачах маршрутизации на транспорте различных классов.
}

%%%%%%%%%%%%%%% To define new section use a command \subsection
% We are defining "Introduction" here
\subsection{Introduction}
%%%%% Paragraph text. To separate paragraphs use empty lines.
Одним из средств экономии ресурсов при транспортировке грузов является применение систем поддержки принятия решений в сфере транспортной логистики. Одной из ключевых функций систем поддержки принятия решений в сфере транспортной логистики является возможность расчета и построения эффективных с точки зрения стоимости объезда маршрутов различного назначения на транспортной сети. Математическая формулировка этой задачи широко известно как задача маршрутизации транспорта. Существует ряд разновидностей ЗМТ с различными дополнительными условиями, которые позволяют учитывать грузоподъемность транспортных средств и другие ограничения для более полного представления деталей реальной действительности.

Данная статья посвящена рассмотрению задачи курьерской маршрутизации, которая является подзадачей ЗМТ. Данная задача состоит в нахождении маршрутов для посещения заданного множества адресов некоторым количеством единиц транспортных средств. Такая проблема наиболее часто встречается среди компаний, выполняющих развозки товара от отправителя к получателю.

Автоматизация процесса планирования маршрутов будет актуальной для:  интернет-магазинов, крупных оптовых компаний (например, для развозки товаров, которые быстро портятся или оптимизации схемы доставки товаров в пункты назначения), фирм с целью организации допоставок в магазины. Практическое применение задачи можно встретить в сфере транспортировки пациентов между корпусами поликлиники\cite{Beaudry, Kergosien}, доставке газет и журналов, доставки топлива в частные дома \cite{Sarak}. Решение рассматриваемой задачи также имеет практическое применение в системах вертолетных перевозок людей между морскими нефтяными платформами \cite{Romero}.

В последнее время все больше людей пользуются услугами курьерских служб.  Обычно заявки формируются днем ранее и
перед диспетчером полный список заявок, которые необходимо выполнить в кратчайшие сроки. Тематическое исследование провел Attanasio(2007) \cite{Attanasio}, в котором описал преимущества использования компьютерных методов перед человеческой диспетчеризацией. В своей работе он рассмотрел примеры работы eCourier Ltd, лондонской компании, которая предлагает курьерские услуги. Их клиентами в основном являются юридические фирмы, финансовые учреждения, рекламные агентства и другие организации, которые заинтересованы в быстрой доставке товаров или оригинальных подписях на документах. После того как все заявки на конкретный момент времени сформированы, необходимо составить такой маршрут, который удовлетворял бы требованиям заказчиков. В зависимости от уровня обслуживания, который был задан клиентом, курьер может объединить или ни его доставку с другими, то есть это означает, что товары могут быть доставлены по очереди, или курьер может выполнить ускоренную доставку.

Компании, которые предлагают курьерские услуги достаточно часто имеют смешанный парк транспортных средств, который состоит из велосипедов, мотоциклов, автомобилей и небольших грузовиков. В зависимости от типа запроса, расположение и временных окон будет выбран тот вид транспортного средства, который сможет выполнить заказ в кратчайшие сроки. Исследования \cite{Attanasio} показали, что использование компьютерных методов, в том числе алгоритмов оптимизации, были очень выгодными для курьерских компаний.

Таким образом, использование автоматизированной системы позволяет улучшить качество обслуживания, время доставки, повысить курьерскую эффективность и сократить расходы на доставку, тем самым обеспечить повышение конкурентного преимущества.

\subsection{Постановка задачи}
Пусть $C,dim(C)=n$ -- множество транспортных средств, $Q,dim(Q)=m$ -- множество запросов клиентов, которые требуется обработать на текущий момент времени.

Предположим, что следующая информация известна о транспортных средствах из множества $C$: \\
$\vec{P}_c$ -- местоположение ТС, $c \in C$;\\
$L_c$ -- грузоподъемность ТС, $c \in C$.

Пусть $S$ -- множество отправителей груза, ($dim(S) = dim(Q) = m$), $R$ -- множество получателей, $dim(R) = dim(Q) = m$. Тогда каждый клиентский запрос $q \in Q$ включает в себя следующую информацию:\\
$s_q$ -- отправитель груза клиента, $s \in S$; \\
$r_q$ -- получатель клиентского груза, $r \in R$; \\
$\vec{P}_{s_q}$ -- Местоположение отправителя;\\
$\vec{P}_{r_q}$ -- Местоположение получателя; \\
$w_q$ -- клиентский груз, который требуется перевезти от отправителя к получателю; \\
$[t_{s}^{q}, t_{s}^{q}+ \Delta{t_{s}^{q}}]$ -- временное окно, в течение которого нужно забрать груз у отправителя;\\ 
$[t_{r}^{q}, t_{r}^{q}+ \Delta{t_{r}^{q}}]$ -- временное окно, в течение которого нужно доставить груз получателю.

Таким образом, каждая заявка может быть представлена в виде кортежа:
\begin{equation}
\forall q \in Q: \exists q = (s_q,r_q, \vec{P}_{s_q}, \vec{P}_{r_q}, w_q, t_{s}^{q}, \Delta{t_{s}^{q}}, t_{r}^{q}, \Delta{t_{r}^{q}})
\end{equation}
Для того, чтобы оценить стоимость перевозки между пунктами назначения, определим функцию стоимости $\Omega$:
\begin{equation}
\forall i,j \in S \cup R: \exists \Omega_{i,j} = \Omega(\vec{P}_i,\vec{P}_j)
\end{equation}

Необходимо построить оптимальные маршруты движения транспортных средств для перевозки грузов от отправителя к получателю для всех клиентских запросов.

@@

\subsection{Дискретная модель}
Задачу маршрутизации можно представить в виде ориентированного графа $G=G(V,E)$. Пусть множество $V=C\cup{S}\cup{R}$ -- узлы графа $G$,  состоящее из элементов множеств транспортных средств $C$, отправителей $S$ и получателей $R$. $E$ -- динамическое множество дуг графа $G$, такое, что:
\begin{equation}
\forall e(X) \in E: e(X) = (v_i,v_j), \exists v_i \in V, v_j \in V/C
\end{equation}

Пусть $X = \{X^k\}^n_{k=1}$ представляет собой последовательность матриц для каждого ТС $k \in C$. Элементы матрицы принимают следующие значения:
\begin{equation}
  x^{(k)}_{i,j} = 
    \begin{cases}
	  1,&\text{АТС $k \in C$ движется с узла $i$ в вершину $j$}\\
	  0,&\text{в противном случае.}
    \end{cases}
\end{equation}
где: $i\in{V/C \cup {k}}, j \in V/C$.

Введем вектор переменных $ \vec{Y}  ^ k(X) $ для каждого транспортного средства  $ k \ in C $. Векторные элементы принимают следующие значения:
\begin{equation}
\vec{y}^{(k)}_{j}(X) = 
\begin{cases}
1,&\text{Запрос $j \in Q$ обслуживается ТС $k \in C$}\\
0,&\text{в противном случае}
\end{cases}
\end{equation}
где: $i\in{V/C \cup {k}}, j \in V/C$.

Пусть $t^k_j(X^{(k)})$ -- время прибытия автомобиля $k \in C$ в пункт назначения $j \in S \cup R$.

Целевая функция принимает следующий вид:
\begin{equation} \label{main_objective}
  F(X) = 
    \sum_{k \in C}
     \sum_{i,j\cup{V}} 
     \Omega_{ij} \cdot x_{ij}^{(k)} 
     \to min
\end{equation}

Наложим ограничения на целевую функцию  (\ref{main_objective}) для обеспечения непрерывности маршрутов:
\begin{align} 
& \sum_{k \in C}\sum_{j \in S \cup R}x^{(k)}_{i,j} \leq 1, 
\forall i \in V \label{main_cond_1}\\
& \sum_{k \in C}\sum_{i \in S \cup R \cup \{k\} } x^{(k)}_{i,j} = 1, 
\forall j \in S \cup R \label{main_cond_2}\\
& \sum_{i \in S \cup R \cup \{k\} } x^{(k)}_{i,\omega} - 
\sum_{j \in S \cup R} x^{(k)}_{\omega,j} \leq 1, 
\forall \omega \in S \cup R,  \forall k \in C \label{main_cond_3}\\
&  \sum_{i \in S \cup R / Z}\sum_{j \in Z } x^{(k)}_{i,j} > 0, 
Z=\{z \in Z: \sum_{j \in S \cup R}x^{(k)}_{j,z}>0 \}  ,\forall k \in C \label{main_cond_4}
\end{align}

Ограничение (\ref{main_cond_1}) запрещает, чтобы у узла графа $G$ было более чем одна выходящая дуга. Ограничение (\ref {main_cond_2}) запрещает узлу иметь более чем одну входную дугу. Ограничение (\ref {main_cond_3}) указывает на то, что у узла количество входных дуг не может быть меньше количества выходных дуг (это ограничение учитывает тот факт, что транспортное средство может покинуть пункт назначения, если только он посетил этот узел). Ограничение (\ref {main_cond_4}) исключает локальные циклы.

Следующие ограничения синхронизируют значения переменных $X$ и $ \Vec{у} $ для каждого запроса $ Q \ в Q $ и запрещают посещение получателя перед отправителем:

\begin{align} 
& x^{(k)}_{s_q} + x^{(k)}_{r_q} = 2 \cdot y^{(k)}_{q}(X), \forall k \in C, q \in Q  \\
& y^{(k)}_{q}(X) \cdot (\tilde{t}^k_{r_q}(X^{(k)})-\tilde{t}^k_{s_q}(X^{(k)})\ge{0}, \forall k \in C, q \in Q
\end{align}

Определим ограничения для учета грузоподъемности транспортных средств и временные окна:
\begin{align} 
& \sum_{j\in{Q}} \omega_j \cdot y_{j}^{k} \leq L_k, \forall{k}\in{C}\\
& t_{s}^{q} \leq \tilde{t}^k_{s_q}(X^{(k)} \leq t_{s}^{q}+ \Delta{t_{s}^{q}}, \forall q \in Q, \label{tws_cond} \\
& t_{r}^{q} \leq \tilde{t}^k_{r_q}(X^{(k)} \leq t_{r}^{q}+ \Delta{t_{r}^{q}}, \forall q \in Q, \label{twr_cond}
\end{align}

\subsection{Описание алгоритма}
Представим множество матриц $\{X^k\}$ в виде вектора, определенного на гиперкубе $E^\eta=\{0.1\}^\eta$, где $\eta=n\cdot(2m+1)\cdot 2m$. Цель задачи состоит в минимизации следующей целевой функции:

\begin{align} 
& F(\vec{u})\to min,\vec{u}\in E^{\eta}
\end{align}

Обозначим через $\delta(\vec{u},\vec{v})$ расстояние Хэмминга между $\vec{u}$ и $\vec{v}$. Через $N_l(\vec{u})$ обозначим окрестность точки $\vec{u}$ дариусом $l$ \cite{vrp_shafeyev}:
\begin{align} 
& N_l(\vec{u})=\{\vec{v} \in E^{\eta}:\delta(\vec{u},\vec{v})\le l \}, l=\bar{1,\eta}
\end{align}

При $ л = \ ETA $ множество $ N_l (\ VEC {и}) $ для любого вектора $ \ VEC {и} $ совпадает с множеством $ E ^ {\ эты} $ и быть в этой окрестности вектора с минимальным значение целевой функции эквивалентна решению исходной задачи. Стандартный алгоритм Локальный алгоритм поиска начинается с выбранного случайным образом вектор $ \ VEC {и ^ 0} $.
На $ I $ шаг алгоритма перехода от текущего вектора в соседней, что минимальное значение целевой функции в окрестности данного вектора:

Когда $l=\eta$, тогда множество $ N_l(\vec{u})$ для любого вектора $\vec{u}$ совпадает с множеством $E^{\eta}$  и поиск в этой окрестности минимума эквивалентно решению исходной задачи. Классический алгоритм локального поиска начинает поиск оптимума с  инициализации вектора $\vec{u^0}$.
На $i$ шаге алгоритма  алгоритма выполняется перемещение текущего вектора в минимум его окрестности:
\begin{align} 
& F(\vec{u}^{i+1})=min\{F(\vec{v}):\vec{v} \in N_l(\vec{u}^i)\}
\end{align}

Алгоритм завершается в локальном оптимуме, когда $F(\vec{x^{i+1}})=F(\vec{x^i})$. В задачах маршрутизации, как правило, имеется большое множество локальных оптимумов и только один из них носит глобальный характер:
\begin{align} 
& F_{opt}=min\{F(\vec{v}):v \in E^{\eta}\}
\end{align}

@@

\subsection{Ініціалізація}
Для того, щоб використовувати алгоритм імітації відпалу необхідно побудувати початкове рішення. Для отримання початкового рішення $\vec{u^0}$ можна скористатися евристичним методом конструювання маршруту. Суть цих методів полягає в наступному. Для кожного автомобіля послідовно будується маршрут шляхом додавання ще не розглянутих клієнтських заявок за заданими правилами. У нашому випадку, для збереження порядку відвідування вершин, під замовленням розуміється пара "відправник-одержувач".

Найвідоміший метод конструювання маршруту був запропонований Соломоном у 1987 році \cite{Cordeau2001,Braysy} і в літературі зустрічається як метод Соломона. Даний алгоритм є досить швидким (обчислювальна складність $O(n^3)$ \cite{Babb}), тому його зручно використовувати для ініціалізації вектора $\vec{u^0}$.

На початку роботи алгоритму відбувається додавання по одному запиту на маршрут у відповідності з тимчасовими обмеженнями. Вибір першої заявки відбувається випадковим чином, або вибирається та, яку необхідно виконати раніше за інших. Далі кожна можлива заявка $u$ з множини нерозглянутих заявок $Q$ (множина $Q' \in Q$) оцінюється на предмет вставити її в початок маршруту, кінець або між двома сусідніми вузлами маршруту. Для вибору місця вставки використовується наступний критерій:
\begin{align} 
	\label{insert_kr}
	c(k,v_i,u,v_{i+1}) = \min =
	\begin{cases}
		\min_{q=2,\dots,n^k}(\Omega_{q-1,u}^k+\Omega_{u}^{k}+\Omega_{u,q}^{k}-\mu \Omega_{q-1,q}^k),\\
		\Omega_{k,u}^k+\Omega_{u,q}^k+\Omega_{u}^k\mu \Omega_{q,q}^k), q=1\\
		\Omega_{q,u}^k+\Omega_{u}^k, q=n^k
	\end{cases}
\end{align} 
де:\\
$k$ -- маршрут транспортного засобу $k\in C$;\\
$n^k$ - кількість вузлів у маршруті;\\
$\mu$ -- довільний параметр, $\mu \ge 0 $;\\
$\Omega_{u}^k=\Omega^k[\vec{P}_{s_q}, \vec{P}_{r_u}]$ -- варіант вставки нової заявки між вже існуючими відправником і одержувачем;\\
$\Omega_{u,q}^k=\Omega^k[\vec{P}_{r_u}, \vec{P}_{q}]$ -- варіант вставки нової заявки на початок маршруту;\\
$\Omega_{q,u}^k=\Omega^k[\vec{P}_{q}, \vec{P}_{s_u}]$ -- варіант вставки нової заявки в кінець маршруту.

Додається та заявка, у якій $c(k,v_i,u,v_{i+1})$ буде мінімальною:
\begin{align} 
	\label{kriterij}
	& u^*:c(k,v_i,u^*,v_{i+1})=\min_{u}[c(k,v_i,u,v_{i+1})]
\end{align} 
В результаті $u^*$ додається до поточного маршруту $k$ на найвигіднішу позицію.
\begin{figure}[h]
	\center{\includegraphics[width=1\linewidth]{images/pict1}}
	\caption{Приклад виконання одного кроку методом Соломона}
	\label{ris:pict1}
\end{figure}

На рис. \ref{ris:pict1} представлений зважений граф, на прикладі якого показано виконання одного кроку методом Соломона:\\
а) вихідний граф, на якому показані варіанти додавання нових заявок на маршрут;\\
b) маршрут, який був побудований методом Соломона після виконання одного кроку.

\textit{Приклад.} Спочатку дана заявка (яка складається з відправника і одержувача). Значення функції вартості $\Omega$ представлені на дугах графа (рис.\ref{ris:pict1}). Необхідно додати нові заявки в маршрут, використовуючи критерії \ref{insert_kr} і \ref{kriterij}. Коефіцієнт $\mu=0.5$.

\textit{Рішення.} Розглянемо процес додавання нових заявок. З двох запропонованих заявок $(\{S_2,R_2\},\{S_3,R_3\})$ виберемо ту, яку потрібно додати в першу чергу. В залежності від місця куди можна додати заявку (початок, кінець, середина) виберемо потрібний критерій. Таким чином, знайдемо оптимальне рішення вставки нової заявки $\{S_2,R_2\}$:\\
$\min[c(k,S_1, \{S_2, R_2\}, R_1), c(k,\{S_2, R_2\}, S_1, R_1), c(k, S_1, R_1, \{S_2, R_2\})]=\min[3+11+4-0.5 \cdot 8; 17+11+8-0.5 \cdot 20; 2+11]=\min[14;26;13]$\\
Знайдемо оптимальне рішення вставки нової заявки $\{S_3,R_3\}$:\\ 
$\min[c(k,S_1, \{S_3, R_3\}, R_1), c(k,\{S_3, R_3\}, S_1, R_1), c(k, S_1, R_1, \{S_3, R_3\})]=\min[6+9+12-0.5 \cdot 8; 10+9+13-0.5 \cdot 20; 5+9]=\min[23;22;14]$.\\
Із знайдених рішень виберемо рішення з найменшим значенням $c(k,v_i,u,v_{i+1})$. У нашому прикладі це $c(k, S_1, R_1, \{S_2, R_2\}=13$.

Таким чином, новий маршрут буде складатися з початкових заявок і нової заявки $\{S_2,R_2\}$, яка буде додана в кінець маршруту.

\textit{Відповідь}: в маршрут $\{S_1,R_1\}$ необхідно додати заявку $\{S_2,R_2\}$, яка буде розміщена в кінці.

На рис. \ref{ris:pict2} представлені варіанти додавання в маршрут нової заявки $\{S_2,R_2\}$:\\
a) варіант додавання заявки $\{S_2,R_2\}$ в кінець маршруту;\\
b) варіант додавання заявки $\{S_2,R_2\}$ в початок маршруту;\\
c) варіант додавання заявки $\{S_2,R_2\}$ між вершинами $\{S_1,R_1\}$.

На рис. \ref{ris:pict3} представлені варіанти додавання в маршрут нової заявки $\{S_3,R_3\}$:\\
a) варіант додавання заявки $\{S_3,R_3\}$ в кінець маршруту;\\
b) варіант додавання заявки $\{S_3,R_3\}$ в початок маршруту;\\
c) варіант додавання заявки $\{S_3,R_3\}$ між вершинами $\{S_1,R_1\}$.

\begin{figure}[t]
	\center{\includegraphics[width=1\linewidth]{images/pict2}}
	\caption{Варіанти додавання в маршрут заявки $\{S_2,R_2\}$}
	\label{ris:pict2}
\end{figure}

\begin{figure}[t]
	\center{\includegraphics[width=1\linewidth]{images/pict3}}
	\caption{Варіанти додавання в маршрут заявки $\{S_3,R_3\}$}
	\label{ris:pict3}
\end{figure}

\clearpage

\subsection{Метод формування околиці}
Під час роботи алгоритму імітації відпалу необхідно проводити перегляд околиці $N_{l}(\vec{u^i})$ поточної точки $\vec{u^i}$.В ході обчислення маршруту в околицю потрапляє досить велика кількість точок. Але в зв'язку з особливостями обмежень нашої моделі багато з цих точок не є рішенням задачі. Тому в зв'язку з особливостями розглянутої дискретної задачі ми використали наступний прийом визначення околиці.

Нехай $x^{k}(v)$ -- індекс вузла $v$, $v \in R \cup S$ в маршруті автотранспортного засобу $k \in C$. Кожна заявка складається з пари $\{S,R\}$, де $S$ -- це відправник, а $R$ -- одержувач. Під час визначення околиці, повинна зберігатися цілісність маршруту, тобто виконуватися обмеження (\ref{main_cond_1} --\ref{main_cond_4}). Над заявками маршруту $Route_i$, $i=1 \dots n$ виконуються операції перестановки. Отримані під час перестановок вектори $\vec{u}$ і буде околицею $N_{l}(\vec{u^i})$. Для перестановки використовувалися операції переміщення і поглинання.

Операція поглинання представлена на рис.\ref{ris:figures}(a), на якому представлено два маршрути $Route_1$ і $Route_2$. Маршрут $Route_1$ складається з вузлів $\{C_1,S_1,R_1,S_2,\\R_2,S_3,R_3\}$, а маршрут $Route_2$ складається з вузлів $\{C_2,S_4,R_4,S_5,R_5,S_6,R_6\}$. Під час виконання операції поглинання заявки $\{S_1,R_1\}$ в маршруті $Route_1$, вставка даної заявки в маршрут $Route_2$ може бути виконана на $n+1$ позицій.

\begin{figure}[h]
	\center{\includegraphics[width=0.75\linewidth]{images/neighboard.png}}
	\caption{Приклад операції a)поглинання та b)переміщення}
	\label{ris:figures}
\end{figure}

Операція переміщення представлена на рис.\ref{ris:figures}(b) На даному рисунку представлений маршрут $Route_1$, який складається з вузлів $\{C_1,S_1,R_1,S_2,R_2,S_3,R_3\}$. Для збереження цілісності маршруту вузол $S_i$ (відправник) має бути відвіданий до $R_i$ (одержувач), виходячи з цього під час операції переміщення, вузол $S_i$ може переміщатися поки буде виконуватися умова $0<x^{new}(S_i)<\widetilde{u}(R_i)$, а вузол $R_i$ може переміщатися поки буде виконуватися умова $n^k \geq x^{new}(R_i)>\widetilde{u}(S_i)$.

@@

As should have been expected, for large values of the parameter a jam occurs in local optimum due to the large number of restrictions of movement in space. Otherwise if you select a too small tabu size, algorithm can still get stuck in a local optimum. In the first computational experiment we made finding the best solutions for different values of the tabu size parameter from interval $h \in [0, 40]$ (pict. \ref{aba:fig2} an pict. \ref{aba:fig3}).
\begin{figure}[h]
	\hfil\includegraphics[height=2.2in]{images/tabuSize_time}\hfil
	\caption
	{
	  The dependence of the quality of solutions on the size of the tabu list when using $ TimeSpent $ termination strategy
	}
	\label{aba:fig3}
\end{figure}

\begin{figure}[h]
	\hfil\includegraphics[height=2.2in]{images/acceptedCountLimit}\hfil
	\caption
	{
	  The dependence of the quality of solutions on the neighborhood size when using $ TimeSpent $ termination strategy
	}
	\label{aba:fig4}
\end{figure}
